Staying Hydrated: The Molecular Journey of Gaseous Sulfur Dioxide to a Water Surface

### 2012-03-28 resubmission/ResponseToReviews.tex

\documentclass{article}

\usepackage {textcomp}
\usepackage {setspace}
\usepackage [pdftex]{graphicx}

% some formatting tags
\usepackage[
top    = 0.5in,
bottom = 1.5in,
left   = 2.50cm,
right  = 2.50cm]{geometry}

\begin{document}

%\doublespacing
Journal: The Journal of Physical Chemistry

Manuscript ID: jp-2011-11804m

Title: "Staying Hydrated: The Molecular Journey of Gaseous Sulfur Dioxide to a

Water Surface"

Author(s): Shamay, Eric; Richmond, Geraldine 
\\ \\
Dear Prof. Jungwirth,

Because of the depth and impact of the reviewer comments about our recently submitted manuscript, the authors have decided to completely re-run the simulations of SO2 on water in order to bring the simulation parameters into line with the reviewer suggestions. This major reworking of the simulations and data resulted in major changes to the data presented, manuscript text, results and conclusions. We have put forth great effort to respond to each of the reviewer comments. Please find our enumerated responses below, matching to each numbered point of the original reviewer comments. We believe that with the changes to the work and the newly run simulations, the new manuscript is more accurate and in line with current simulation standards for aqueous systems. We hope you will further consider our manuscript for publication in the journal.
\\ \\

--Eric Shamay \\ \\

\textbf{Reviewer: 1} \\

1,2,3. The authors have taken into consideration the entirety of all the reviewer suggestions regarding functional and basis sets, cutoffs, timesteps, and temperature that affect the simulations. The changes to the simulation parameters are reflected in the text, and meet the standards suggested in the reviews. All the simulations have been rerun, data has been re-analyzed, and the results are all presented from the new simulations.
\\
% 1. The computational parameters chosen as well as the authorsʼ claim that “these computational parameters were verified to yield a reasonable description of bulk room temperature water …” runs counter to published results.

@@

%2. Given that the system contains water and thus the highest vibrational modes are ~3700 cm-1, it is hard to fathom a timestep of “1 fs” can be used to accurately integrate the equation of motions without energy drift. If this is truly the case, it might behoove the authors to provide some supplemental materials showing the conserved quantity as a function of time for both the NVT and NVE simulations provided the system is not glassy and can sample all the possible conformations.

%3. The author stated that DZVP basis set was chosen, it was never explained why the authors believe that such a basis set is adequate. In fact, work by VandeVondele et al shows that at least a TZV2P basis set with a cutoff of 280 Ry was necessary in order to obtain forces comparable to other more established codes such as CPMD. In addition, works by Siepmann et al on vapor liquid coexistence curves hints that the standard TZV2P basis set with a cutoff of 280 Ry might still be inadequate. Last, what was the convergence criteria used for the SCF?

4. The cold system temperature was set to 210K in the resimulated systems. This much larger separation of temperatures is more in line with the reviewer's concerns regarding temperature separation. Furthermore, included with this revised submission are plots of the temperature and the conserved quantity for each timestep for both low and high temperature systems. As is evident, the temperature increases until reaching the setpoint temperature soon after the start of the simulations. The plateau region of the temperature extends for the majority of the simulations. Lastly, it is clear that the separation of temperatures persists for the entire simulation time, confirming that the systems are representing temperatures that are distinct. The authors believe that this is evidence that the parameter set chosen for simulation of these small water systems are sufficient to recreate the expected temperature conditions, even when simulating in the NVE ensemble.
\\
%4. Following recent literatures on simulations of liquid water using potentials derived from density functional theory (albeit some with different codes as well as different methods), one firm conclusions can be made is that water near ambient conditions are more glassy than liquid. One way to overcome this is through massive thermostating which in addition also ensures equipartition of the kinetic energy in all degrees of freedom. This raises two issues.  a. First, as the production run was obtained in the microcanonical (NVE) ensemble and the two target temperature is only separated by 27 degrees, can the authors show that temperature for the subsystem of interest is really the targeted temperature thus fulfilling the equipartition theorem.  b. And can the authors show that indeed, the water part of the simulated system is behaving as a liquid and not conformationally trap in a glassy state?

5. The temperature coupling constant was added to the text, and the typo was revised.
\\
%5. “Nose” should be “Nosé”. In addition, the coupling frequency, when not chosen appropriately can cause energy drifts, was not specified.

6. The text has been updated to reflect the correct equilibration period of 1 ps.
\\
%6. The authors indicate the system was equilibrated for “1 ns”. This just does not seem correct unless the equilibration was done using a cheaper potential other than DFT.

7. Both RDFs converge to 1.0 because of the normalization routine employed in the calculations. This is typical of RDF calculations - normalized to unity - and the authors prefer this normalization as it aids in comparison to the other RDFs published for similar systems (see Baer et al. cited in the text)
	\\
%7. From the RDF, it can be seen that both hot and cold simulations converge to 1.0 as distance increases. But if the sulfur dioxide gas molecule is at the liquid-vapor interface, shouldnʼt this value approach something more like 0.5 instead of 1.0?

8. The inclusion of dispersion corrections were not explored for this simulation study. The goal of the project was to monitor specific bonding behaviors of SO2 using previously established methods, not as a study of the results from various constraints or changes to simulation parameters. The authors believe that additional discussion on this topic would distract from the narrative presented.
\\
%8. Has the authors explore the ramifications of including dispersion correction in their calculation which might play a large in such a system?

\textbf{Reviewer: 3} \\

1. Because the re-simulated systems were done using B3LYP functional, the authors believe a benchmark comparison is no longer necessary. Small cluster studies of so2 with several waters have been performed using high-accuracy basis and functionals previously, and we do not set out to reproduce that work here, but rather to build on it. The authors believe that the current work with the new simulation parameters achieves a level of accuracy suggested by the reviewers, but that lengthy discussion of this fact is not requisite in this particular simulation study.
\\
%1.	The predominant stabilization interaction between and SO2 molecule and a water molecule is due to dipoles. This essentially places the sulpher close to the water oxygen. This simple dipole interaction reasoning is consistent with everything the authors find. However, there is a component of this interaction that arises due to van der Waal’s, due to the sulpher lone pairs. In this regards BLYP has been used here, which is known not to capture such subtle interactions. Considering this, I would have liked to see at least a set of benchmarks for the functional and basis used in this work, with post Hartree-Fock MP2 and B3LYP used for comparison. The benchmark system could have involved a simple cluster of SO2 with a few water molecules. The absence of this comparison makes the results weak.

2. After resimulating the systems the resulting relative populations of unbound SO2 has changed. The more intuitive stronger binding of SO2 at the cold temperature results from the higher accuracy of the functional and basis sets used in the new simulations. The new results are also more in line with the author's experimental results from the previous publications cited in the text.
\\
%2. The larger concentration of unbound SO2 at a lower temperature is counter-intuitive. (Fig. 4) One would have expected the opposite trend and no physical reasoning is provided to explain this result.

3. Several previous simulation studies have discussed the structure of water under similar simulated conditions. The focus of the work is on the structure of SO2 and its behavior near a water cluster surface. Thus, the authors have made the conscious decision to avoid lengthy discussion of the surrounding water's structure.
\\
%3.	The structure of the surrounding water molecules is never discussed.

4. Reference to the Car-Parrinello method has been added to the text.
\\
%4.	It is stated that Quickstep is used. Assume this implies that the Car-Parrinello method is used, but it is not stated in the paper.

5. The new systems were simulated with a smaller number of water molecules and are referred to as a small water cluster. The effects of using few water are not discussed in detail as the number of waters used is representative of many small-cluster MD and DFT studies. The authors believe that the number of waters is in line with several previous studies of molecules bound to small water clusters, and further discussion is not warranted as part of the narrative being presented.
\\
%5.	The system includes 36 water molecules in a box. The effect of this restriction is important to discuss.

6. K-point integration was not used in the integration, as this was not addressed in similar previous studies. Also, the CP2K simulation package does not allow for this type of calculation. The authors do not believe this is detrmiental to the accuracy of the calculations and further discussion is not pertinent to this study.
\\
%6.	Was k-point integration used in the calculation? If so how many. If only gamma-point was used, again there needs to be a discussion on the effectiveness of the calculation.

7. In the new simulations, the timestep has been reduced to 0.5 fs. The text has been updated to reflect this.
\\
%7.	The time-step of 1fs appears to be too large for a Car-Parrinello simulation. Since NVE simulations are used, I would be interested in knowing how well the total system energy is conserved (which determines the extent to which the dynamics simulations are accurate).

8. The authors thank the reviewer for pointing out this previous study using graph theoretical techniques (Oktay Sinanoglu J. Am. Chem. Soc., 1975, 97 (9), pp 2309–2320). That study pertains to synthetic or catalytic pathways of organic systems, but can be generalized to very convoluted and complicated reaction pathways leading to very intricate graphical networks. The systems investigated in that work are much more complicated, and the graphs are designed to convey a great deal more information than in the present text. In our work we wish to investigate only the transitions between different coordination or cyclic-hydrate states, and are not interested in multi-reactant processes. A citation to the work has been added to the text.
\\
% 8.	With respect to previous work on molecular computations using graphs, I would like to bring to the author’s attention the following work by O. Sinanoglu, which to my knowledge is one of the first such attempts: Oktay Sinanoglu J. Am. Chem. Soc., 1975, 97 (9), pp 2309–2320.

\end{document}

### 2012-03-28 resubmission/Article.tex
New RDF images
Finishing up with the resubmission draft, and the responses to the reviewers. Added plots of the temp and conserved quantities for two representative plots at cold/hot to submit with the response to reviews.

To detect cyclic structures in a graph a depth-first or breadth-first search (DFS and BFS, respectively) may be used.\cite{Knuth1997, Cormen 2001} A graph search is a recursive algorithm of queuing nodes and all neighboring nodes while performing a specified procedure on each visited node. This is easily performed on adjacency list or connectivity matrix data structures, iterating through nodes (i.e. atoms) of interest in the graph as starting points of the search. In graph search terminology, all nodes are colored during graph traversal to distinguish unvisited nodes (white), queued nodes (gray), and visited nodes (black). Performing a BFS on a graph, cyclic structures are detected any time a ``gray target'' is encountered when queuing adjacent neighbors of a node. A benefit of BFS is the ability to determine the smallest cyclic structure containing a given node. In the case of \suldiox~hydrate structures, beginning the BFS with the \suldiox-sulfur as the starting, or root node for the search, will discover cyclic bonding structures in order of size. Here we are only concerned with the smallest cyclic structure involving those waters in the first and second hydration shells around the \suldiox. Furthermore, it is possible to reconstruct a cycle's structure by finding its size (number of contributing atoms), and the number of unique waters in the cycle. This allows us to distinguish between various types of cyclic structures encountered.

@@

\begin{figure}[h!]
	\begin{center}
		\includegraphics[scale=1.0]{cycle-types-small.png}
		\caption{\suldiox~in various cyclic structures encountered during MD simulations with water. The cartoons show the five types of cyclic structures, numbered for reference. The cyclic structures may involve any number of waters, but here each structure is shown with three waters. Type \Rmnum{3} is formed by a \suldiox~with the ``SO'' bonding coordination, which is the most dominant bonding coordination encountered. Each hydrate structure is labeled with its corresponding bonding coordination type using the nomenclature described in the text.}
		\label{fig:cyclic-structures}
	\end{center}
\end{figure}

@@

On-the-fly ab initio molecular dynamics simulations were performed with the QUICKSTEP package, which is an implementation of the Gaussian plane wave method using the Kohn-Sham formulation of density functional theory (DFT).\cite{VandeVondele2005} The Kohn-Sham orbitals are expanded using a linear combination of atom-centered Gaussian-type orbital functions. The electronic charge density was described using an auxiliary basis set of plane waves. Energies and forces from on-the-fly simulation sampling of the Born-Oppenheimer surface were calculated for each MD step using the Gaussian TZV2P basis set with a cutoff of 280 Ry, the exchange-correlation functional of Becke, Lee, Yang, and Parr (B3LYP),\cite{LEE1988} and the atomic pseudo-potentials of the Goedecker, Teter, and Hutter type.\cite{Goedecker1996} A simulation timestep of 0.5 fs was used, with a Nos\'{e}-Hoover thermostat set at 210K and 300K for the ``cold'' and ``hot'' simulations, respectively. %These computational parameters were verified to yield a reasonable description of bulk room temperature water when simulating a neat-water system.

Initially, 20 equilibrated boxes of side-lengths 8.545\angs, with 24 randomly packed water molecules were used. Ten of the boxes were used for each of the cold and hot simulations. A sulfur dioxide molecule was randomly placed onto the surface within 2.5\angs~of a water molecule centrally located above the waters in the z-axis. A copy of the initial system cubes were then expanded along one axis (z-axis) to 30\angs. The system energy was minimized through a geometry optimization. Subsequently, the system was equilibrated for 1 ps in canonical ensemble (NVT) conditions. Periodic boundaries were set on the two short axes to form an infinite slab. The equilibrated systems were then simulated for a further 20 ps in the microcanonical ensemble (NVE), with trajectory snapshots recorded every 0.5 fs. The initial 1 ps equilibration trajectory was not included in the final analysis. This simulation process resulted in 40,000 time steps of system trajectory for analysis in each of the hot and cold replicas of the system, for a total of 400,000 timesteps at each temperature.

@@

The bonding coordination of each \suldiox~was determined at each timestep of the simulations. Figure \ref{fig:bonding-coordinations} shows the distribution of bonding coordinations of the surface \suldiox, as a percentage of all bonding coordinations encountered for both the cold (blue) and hot (red) trajectories. A first visual inspection reveals several trends. At both temperatures the \suldiox~spends much of the time unbound without any bonding to the surface waters. The cluster of the three most populated coordinations is comprised of the ``S'', ``SO'', and ``SOO''. This grouping represents the coordinations with a single bond through the sulfur, and up to two interactions through the \suldiox-oxygens. In this coordination group the ``SO'' is the most populous, but in no way dominates the bonding types as the difference in populations is less than 5\% within each temperature. The second and third most populated bonding coordinations (excluding the unbound state) are the ``S'' and ``SOO'', however their distributions differ between temperatures. In cold simulations the ``S'' and ``SOO'' coordinations occur nearly equally. The hotter temperature simulation shifts the distribution such that the ``S'' occurs approximately 4\% less frequently than in the cold, and the ``SOO'' occurs nearly 8\% less often. The distribution of bonding coordinations in the hot temperature has a clear first and second most frequent coordination: ``SO'' and ``S'', respectively.

@@

	Several conclusions about the bonding behavior of \suldiox~to surface waters stem from this distribution of coordinations. Clearly the hot \suldiox~spends more time than the cold \suldiox, 45\% versus 23\%, respectively, completely unbound from the surface waters. This does not necessarily imply a complete desorption into the gas phase, but likely only a brief sojourn away from the surface waters, with all interactions and bond lengths longer than the cutoff criteria used for the analysis. The unbound coordination data is also supported by our recent spectroscopic experiments that resulted in increased \suldiox~adsorption at colder temperatures.\cite{Ota2011} Furthermore, the most frequently occurring bonding coordinations are ``S'',``SO'', and ``SOO'', with ``SO'' being the most populated at both temperatures. Baer et al. also concluded that these three coordinations were the most frequent for a room temperature simulation, and specifically identified the ``SO'' and ``SOO'' as most common in their study.\cite{Baer2010}

	Looking closer at the coordination types it is notable that coordinations lacking any sulfur interactions (e.g. ``O'', ``OO'', etc.) represent the least frequently formed. A bonding coordination with at least a single sulfur interaction is clearly favored over \suldiox~``oxygen-only'' bonding to waters. In our previous classical simulations of \suldiox~on water we concluded that during adsorption and throughout the interface, the \suldiox~orients so that its sulfur tends to the \wat~bulk side of the interface.\cite{Shamay2011} The coordination distributions here support the idea that binding through the sulfur is preferable, to the extent that a non-sulfur coordination is not often formed during the course of all our simulations.

% Baer's assymetric oxygen binding
Baer et al performed this coordination analysis for their single-temperature study, but discriminated between \suldiox~binding through the two different oxygens. They concluded that there is asymmetric hydrogen bonding through the \suldiox-oxygens, with one oxygen binding more often than the other. This is supported by the findings here where the double oxygen coordinations (e.g. ``OO'', ``SOO'', etc.) represent a lower percentage of the coordinations than the single oxygen counterparts (e.g. ``O'', ``SO'', etc.). Furthermore, a triple-oxygen coordination (e.g. ``OOO'', ``SOOO'', etc.) is very rarely encountered. Three \suldiox-oxygen bonds only form if both \suldiox-oxygens are interacting with water hydrogens. Our finding that triple-oxygen coordinations rarely form, and that double-oxygen coordination are less frequent than the single-oxygen counterparts, complements Baer et al's conclusion about the asymmetry in the oxygen interactions.

Having established the preference for an interaction through the \suldiox-sulfur atom, we look to the right side of Figure \ref{fig:bonding-coordinations} at the double-sulfur coordinations (e.g. ``SS'', ``SSO'', ``SSOO'', etc.). From the data it is clear that single-sulfur coordinations are overwhelmingly preferred over double-sulfur ones. Adding a third oxygen atom is also unfavorable as the ``OOO'', ``SOOO'', and ``SSOOO'' together represent less than 3\% of the trajectories, and the comparison between ``SOO'' to ``SOOO'' shows a very large decrease in occurrences.

A picture can now be formed of a typical \suldiox~molecule adsorbed to a water surface across both temperatures in this study. Often the \suldiox~will remain unbound from the interface, nearly twice as often when the temperature is hot. The \suldiox~will mostly have at least one interaction to neighboring waters through the sulfur, and will then bond asymmetrically through one of the oxygens either once or twice to water hydrogens. The \suldiox-oxygen bonds will form and break repeatedly throughout a trajectory, and overall the most dominant bonding coordination will be the ``SO'' bonding arrangement.

@@

The binding behavior of the \suldiox~is altered by changing the temperature of the system, as evidenced in the shift in bonding coordination populations of Figure \ref{fig:bonding-coordinations} from cold to hot. In the cold temperature, the bound coordinations are overall more populated than when hot. Conversely, the hot temperature system is much more often unbound (nearly half of the simulated time) than when the temperature is cold. As shown, the cold \suldiox~spends nearly half as much time unbound as the hot \suldiox, with most of the unbound population in the cold simulations shifting to coordinations with single-sulfur bonds.

@@

Figure \ref{fig:rdf} shows the radial distribution functions (RDF) of \suldiox~to water atoms for both cold (blue) and hot (red) temperatures. The S-O$_{H_2O}$ RDFs are nearly equal except for a slightly taller first peak in the cold system. Along with the slightly larger population of the cold ``S'' and ``SO'' coordinations in the cold surface, the RDF indicates that since bonding occurs frequently through the sulfur, the cold \suldiox-sulfur interacts more closely with the surface waters. In the hot systems, the bonding coordinations with two oxygen bonds (e.g. ``SOO'', etc.) occur more frequently than in the cold system. This additional bonding through a second oxygen interaction may slightly shift neighboring hydrating waters away from the \suldiox-sulfur towards the oxygen end of the molecule. This suggests that the cold \suldiox~bonds closer to the surface waters through its sulfur atom (as shown in the S-O$_{H_2O}$ RDF) and favors the more ``sulfur-centric'' bonding coordinations (e.g. ``S'', ``SO''). The increased temperature of the hot system allows the \suldiox~to bond more extensively through its oxygens to the ``SOO'' coordination. The greater interactions through the oxygens, and higher bonding coordination, may pull the \suldiox~further into the water interface and then allow for increased bonding through the sulfur, up to double-sulfur coordinations (e.g. ``SS'', ``SSO'', etc.).

\begin{figure}[h!]
	\begin{center}
		\includegraphics[scale=1.0]{rdf-small.png}
		\caption{Radial distribution functions (RDF) show the subtle change in how the hydrating waters around \suldiox~reposition as the temperature is increased from cold (blue) to hot (green). Shown are the two correlations between sulfur and water oxygens (left) and the \suldiox-oxygen and water hydrogens (right).}
		\label{fig:rdf}
	\end{center}
\end{figure}

@@

\begin{figure}[h!]
	\begin{center}
		\includegraphics[scale=1.0]{coordination-transitions.png}
		\caption{The bonding coordinations of \suldiox~on a water surface rapidly change as bonds break and form throughout the MD trajectories. Depicted here is a graph of the bonding coordinations, shown as colored nodes of varying size. Larger and more darkly colored coordination nodes are those more often encountered during MD (see Figure \ref{fig:bonding-coordinations}). The directed edges between the nodes represent the number of times the coordination transition occurred. Thicker lines correspond to a greater number of transitions between coordinations. The number of times a given transition occurred is labeled to the right of the corresponding line.}
		\label{fig:coordination-transitions}
	\end{center}
\end{figure}

As expected for the more populated coordinations, there are more transitions between larger nodes in Figure \ref{fig:coordination-transitions} (left) than transitions to less populated coordinations. Insights to the bonding process are made clearer from these graphs. In the cold system graph of Figure \ref{fig:coordination-transitions}, we find that the majority of transitions are between the unbound and ``O'' nodes, and between the single-sulfur coordinations (e.g. ``S-SO'', ``SO-SOO'', etc.) The number of transitions within this group of bonding coordinations is an order of magnitude larger than any other transition. This indicates that while the \suldiox~is bound in any of the most populated bonding coordinations, it is actively binding and unbinding the oxygens to form the other coordinations. Clearly, the \suldiox~is rather active and constantly forming and breaking bonds through its oxygens.

In the graph of the hot system in Figure \ref{fig:coordination-transitions} (right), the transition frequencies follow the same trend as in the cold system, increasing with adjacent node size. One very surprising result is in the transition from ``SOO-SOOO''. This transition frequency does not follow from the adjacent node sizes, as the ``SOOO'' node represents less than 5\% of the bonding coordinations. This is indicative of a very rapid cycle of forming and breaking of bonds to the second \suldiox-oxygen. As noted earlier, it is likely that ``SOO'' coordinated \suldiox, asymmetrically binding twice through a single oxygen, is being pulled further into the water interface. It is likely more surrounded by waters, and in the hot system it can more easily form a brief third hydrogen bond to a water through the second \suldiox-oxygen. Because the triple-oxygen coordination is not as favorable, it quickly breaks the bond and the \suldiox~returns to the ``SOO'' coordination.

>>>>

To detect cyclic structures in a graph a depth-first or breadth-first search (DFS and BFS, respectively) may be used.\cite{Knuth1997, Cormen 2001} A graph search is a recursive algorithm of queueing nodes and all neighboring nodes while performing a specified procedure on each visited node. This is easily performed on adjacency list or connectivity matrix data structures, iterating through nodes (i.e. atoms) of interest in the graph as starting points of the search. In graph search terminology, all nodes are colored during graph traversal to distinguish unvisited nodes (white), queued nodes (gray), and visited nodes (black). Performing a BFS on a graph, cyclic structures are detected any time a ``gray target'' is encountered when queueing adjacent neighbors of a node. A benefit of BFS is the ability to determine the smallest cyclic structure containing a given node. In the case of \suldiox~hydrate structures, beginning the BFS with the \suldiox-sulfur as the starting, or root node for the search, will discover cyclic bonding structures in order of size. Here we are only concerned with the smallest cyclic structure involving those waters in the first and second hydration shells around the \suldiox. Furthermore, it is possible to reconstruct a cycle's structure by finding its size (number of contributing atoms), and the number of unique waters in the cycle. This allows us to distinguish between various types of cyclic structures encountered.

@@

\begin{figure}[h!]
	\begin{center}
		\includegraphics[scale=1.0]{cycle-types-small.png}
		\caption{\suldiox~in various cyclic structures encountered during MD simulations with water. The cartoons show the five types of cyclic structures, numbered for reference. The cyclic structures may involve any number of waters, but here each structure is shown with three waters. Type \Rmnum{3} is formed by a \suldiox~with the ``SO'' bonding coordination, which is the most dominant bonding coordination encountered. Each hydrate structure is labelled with its corresponding bonding coordination type using the nomenclature described in the text.}
		\label{fig:cyclic-structures}
	\end{center}
\end{figure}

@@

On-the-fly ab initio molecular dynamics simulations were performed with the QUICKSTEP package (Car-Parrinello method of MD), which is an implementation of the Gaussian plane wave method using the Kohn-Sham formulation of density functional theory (DFT).\cite{VandeVondele2005} The Kohn-Sham orbitals are expanded using a linear combination of atom-centered Gaussian-type orbital functions. The electronic charge density was described using an auxiliary basis set of plane waves. Energies and forces from on-the-fly simulation sampling of the Born-Oppenheimer surface were calculated for each MD step using the Gaussian TZV2P basis set with a cutoff of 280 Ry, the exchange-correlation functional of Becke, Lee, Yang, and Parr (B3LYP),\cite{LEE1988} and the atomic pseudo-potentials of the Goedecker, Teter, and Hutter type.\cite{Goedecker1996} A simulation timestep of 0.5 fs was used, with a Nos\'{e}-Hoover thermostat set at 210K and 300K for the ``cold'' and ``hot'' simulations, respectively, using a coupling constant of 1000 fs. %These computational parameters were verified to yield a reasonable description of bulk room temperature water when simulating a neat-water system.

Initially, 20 equilibrated boxes of side-lengths 8.9545\angs, with 24 randomly packed water molecules were used. Ten of the boxes were used for each of the cold and hot simulations. A sulfur dioxide molecule was randomly placed onto the surface within 2.5\angs~of a water molecule centrally located above the waters in the z-axis. A copy of the initial system cubes were then expanded along one axis (z-axis) to 30\angs. The system energy was minimized through a geometry optimization. Subsequently, the system was equilibrated for 1 ps in canonical ensemble (NVT) conditions. Periodic boundaries were set on the two short axes to form an infinite slab. The equilibrated systems were then simulated for a further 20 ps in the microcanonical ensemble (NVE), with trajectory snapshots recorded every 0.5 fs. The initial 1 ps equilibration trajectory was not included in the final analysis. This simulation process resulted in 40,000 time steps of system trajectory for analysis in each of the hot and cold replicas of the system, for a total of 400,000 timesteps at each temperature.

@@

The bonding coordination of each \suldiox~was determined at each timestep of the simulations. Figure \ref{fig:bonding-coordinations} shows the distribution of bonding coordinations of the surface \suldiox, as a percentage of all bonding coordinations encountered for both the cold (blue) and hot (red) trajectories. A first visual inspection reveals several trends. At both temperatures the \suldiox~spends much of the time unbound without any bonding to the surface waters. The cluster of the three most populated coordinations is comprised of the ``S'', ``SO'', and ``SOO'' bonding types. This grouping represents the coordinations with a single bond through the sulfur, and up to two interactions through the \suldiox-oxygens. In this coordination group the ``SO'' is the most populous, but in no way dominates the bonding types as the difference in populations is less than 2\% and 5\% within the cold and hot simulations, respectively. The second and third most populated bonding coordinations (excluding the unbound state) are the ``S'' and ``SOO'', however their distributions differ between temperatures. In cold simulations the ``S'' and ``SOO'' coordinations occur nearly equally (18.8\% and 18.1\%, respectively). The hotter temperature simulation shifts the distribution such that the ``S'' occurs 4.6\% less frequently than in the cold, and the ``SOO'' occurs 8.6\% less often. The distribution of bonded coordinations (those excluding the ``unbound'' state) in the hot temperature has a clear first and second most frequent coordination: ``SO'' and ``S'', respectively.

@@

	Several conclusions about the bonding behavior of \suldiox~to surface waters stem from this distribution of coordinations. Clearly the hot \suldiox~spends more time than the cold \suldiox, 45.8\% versus 22.8\%, respectively, completely unbound from the surface waters. This does not necessarily imply a complete desorption into the gas phase, but likely only a brief sojourn away from the surface waters, with all interactions and bond lengths longer than the cutoff criteria used for the analysis. The unbound coordination data is also supported by our recent spectroscopic experiments that resulted in increased \suldiox~surface binding at colder temperatures.\cite{Ota2011} Furthermore, the most frequently occurring bonding coordinations are ``S'',``SO'', and ``SOO'', with ``SO'' being the most populated at both temperatures. Baer et al. also concluded that these three coordinations were the most frequent for a room temperature simulation, and specifically identified the ``SO'' and ``SOO'' as most common in their study.\cite{Baer2010}

	Looking closer at the coordination types it is notable that coordinations lacking any sulfur interactions (e.g. ``O'', ``OO'', etc.) form infrequently (approximately 6\% and 8\% of the total hot and cold bonding coordinations, respectively). A bonding coordination with at least a single sulfur interaction is clearly favored over \suldiox~``oxygen-only'' bonding to waters. In our previous classical simulations of \suldiox~on water we concluded that during adsorption and throughout the interface, the \suldiox~orients so that its sulfur tends to the \wat~bulk side of the interface.\cite{Shamay2011} The coordination distributions here support the idea that binding through the sulfur is preferable, to the extent that a non-sulfur coordination is not often formed during the course of all our simulations.

% Baer's assymetric oxygen binding
Baer et al. performed this coordination analysis for their single-temperature study, but discriminated between \suldiox~binding through the two different oxygens. They concluded that there is asymmetric hydrogen bonding through the \suldiox-oxygens, with one oxygen binding more often than the other. This is supported by the findings here where the double oxygen coordinations (e.g. ``OO'', ``SOO'', etc.) represent a lower percentage of the coordinations than the single oxygen counterparts (e.g. ``O'', ``SO'', etc.). Furthermore, a triple-oxygen coordination (e.g. ``OOO'', ``SOOO'', etc.) is very rarely encountered. Three \suldiox-oxygen bonds only form if both \suldiox-oxygens are interacting with water hydrogens. Our finding that triple-oxygen coordinations rarely form, and that double-oxygen coordination are less frequent than the single-oxygen counterparts, complements Baer et al's conclusion about the asymmetry in the oxygen interactions.

Having established the preference for an interaction through the \suldiox-sulfur atom, we look to the right side of Figure \ref{fig:bonding-coordinations} at the double-sulfur coordinations (e.g. ``SS'', ``SSO'', ``SSOO'', etc.). From the data it is clear that single-sulfur coordinations are overwhelmingly preferred over double-sulfur ones. Adding a third oxygen atom is also unfavorable as the ``OOO'', ``SOOO'', and ``SSOOO'' combined represent less than 4\% of the trajectories, and the comparison between ``SOO'' to ``SOOO'' shows a very large decrease in occurrences.

A picture can now be formed of a typical \suldiox~molecule adsorbed to a water surface across both temperatures in this study. Often the \suldiox~will remain unbound from the interface, more than twice as often when the temperature is hot. The \suldiox~will form at least one interaction with neighboring waters through the sulfur, and will then bond asymmetrically through one of the oxygens either once or twice to water hydrogens. The \suldiox-oxygen bonds will form and break repeatedly throughout a trajectory, and overall the most dominant bonding coordination will be the ``SO'' bonding arrangement.

@@

The binding behavior of the \suldiox~is altered by changing the temperature of the system, as evidenced in the shift in bonding coordination populations of Figure \ref{fig:bonding-coordinations} from cold to hot. In the cold temperature, the bound coordinations are overall more populated than when hot. Conversely, the hot temperature system is much more often unbound (nearly half of the simulated time) than when the temperature is cold. A decrease in temperature causes a clear change to the surface binding behavior of \suldiox. At colder temperatures a \suldiox~molecule will bind more strongly to a water surface than at higher temperatures.

@@

Figure \ref{fig:rdf} shows the radial distribution functions (RDF) of \suldiox~to water atoms for both cold (blue) and hot (red) temperatures. The S-O$_{H_2O}$ RDFs are overall very similar. A difference between the cold and hot simulations appears in the first peak of both RDFs. The cold RDFs have a slightly taller and narrower first peak than the hot counterpart, and the peak is located at a slightly shorter distance. Along with the slightly larger population of the cold ``S'' and ``SO'' coordinations in the cold surface, the RDF indicates that since bonding occurs frequently through the sulfur, the cold \suldiox-sulfur interacts more closely with the surface water oxygens than in the hot waters. This suggests that the cold \suldiox~favors the more ``sulfur-centric'' bonding coordinations (e.g. ``S'', ``SO'', ``SOO''). The broader first peak of the hot RDFs likely indicates that the sulfur atoms have a slightly larger array of bonding environments to water oxygens. Similarly, the \suldiox-oxygen displays the same difference in the first peak on the left plot of Figure \ref{fig:rdf}.

\begin{figure}[h!]
	\begin{center}
		\includegraphics[scale=1.0]{rdf-small.png}
		\caption{Radial distribution functions (RDF) show the subtle change in how the hydrating waters around \suldiox~reposition as the temperature is increased from cold (blue) to hot (red). Shown are the two correlations between sulfur and water oxygens (right) and the \suldiox-oxygen and water hydrogens (left).}
		\label{fig:rdf}
	\end{center}
\end{figure}

@@

\begin{figure}[h!]
	\begin{center}
		\includegraphics[scale=1.0]{coordination-transitions.png}
		\caption{The bonding coordinations of \suldiox~on a water surface rapidly change as bonds break and form throughout the MD trajectories. Depicted here is a graph of the bonding coordinations, shown as colored nodes of varying size. Larger and more darkly colored coordination nodes are those more often encountered during MD (see Figure \ref{fig:bonding-coordinations}). The directed edges between the nodes represent the number of times the coordination transition occurred. Thicker lines correspond to a greater number of transitions between coordinations. The number of times a given transition occurred is labelled to the right of the corresponding line.}
		\label{fig:coordination-transitions}
	\end{center}
\end{figure}

As expected for the more populated coordinations, there are generally more transitions between larger nodes in Figure \ref{fig:coordination-transitions} than transitions to less populated coordinations. Insights to the bonding process are made clearer from these graphs. In the cold system graph of Figure \ref{fig:coordination-transitions}, we find that the majority of transitions are between the unbound and ``O'' nodes, and between the single-sulfur coordinations (e.g. ``S-SO'', ``SO-SOO'', etc.) The number of transitions within these groups of bonding coordinations is an order of magnitude larger than any other transition. This indicates that while the \suldiox~is bound in any of the most populated bonding coordinations, it is actively binding and unbinding the oxygens to form the other coordinations, typically those with a single sulfur interaction. Clearly, the \suldiox-sulfur bond to water oxygens is stronger than its oxygen bonds to water-hydrogens. The \suldiox-oxygens are actively and constantly forming and breaking bonds to neighboring waters. In the graph of the hot system in Figure \ref{fig:coordination-transitions} (right), the transition frequencies follow the same trend as in the cold system, increasing with adjacent node size. %One very surprising result is in the transition from ``SOO-SOOO''. This transition frequency does not follow from the adjacent node sizes, as the ``SOOO'' node represents less than 5\% of the bonding coordinations. This is indicative of a very rapid cycle of forming and breaking of bonds to the second \suldiox-oxygen. As noted earlier, it is likely that ``SOO'' coordinated \suldiox, asymmetrically binding twice through a single oxygen, is being pulled further into the water interface. It is likely more surrounded by waters, and in the hot system it can more easily form a brief third hydrogen bond to a water through the second \suldiox-oxygen. Because the triple-oxygen coordination is not as favorable, it quickly breaks the bond and the \suldiox~returns to the ``SOO'' coordination.

Given the information we have about the frequencies of bonding coordination transitions, it is possible to draw a likely route of adsorption beginning with an unbound \suldiox. From the unbound coordination, the \suldiox~can bind to waters either through the sulfur or an oxygen to enter the ``S'' or ``O'' coordinations, respectively. At both temperatures we find that the coordination transition in Figure \ref{fig:coordination-transitions} from unbound to ``O'' occurs over three times more than the transition from unbound to ``S''. We find two possibilities that may explain this difference. The single H-bonding of the ``O'' coordination may form more easily, but also break quickly after formation, accounting for the higher transition frequency. Otherwise, the ``O'' coordination may be the first step in adsorption of an unbound \suldiox, where a subsequent addition of an \suldiox-sulfur interaction to a water oxygen would lead to a transition to the most frequent coordination, ``SO''. In the latter case, any adsorption of \suldiox~proceeds through an oxygen binding, accounting for the increased unbound-``O'' transition frequency, but because it is likely a less-stable coordination, its population is less than the equivalent ``S'' or ``SO'' configuration.
